---
title: "PaperRush Feature Inventory EDA: Weeklies vs SIP"
author: "PaperRush analysis"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
params:
  input_dir: "input"
  output_subdir: "output/feature_inventory_eda"
  row_sample_n: 250000
  top_n_plot: 20
  top_n_preview: 50
  max_interactive_rows: 5000
  include_demographics: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 7
)

required_packages <- c(
  "data.table",
  "dplyr",
  "DT",
  "ggplot2",
  "knitr",
  "skimr",
  "scales",
  "stringr",
  "tidyr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Missing required R packages: ",
      paste(missing_packages, collapse = ", "),
      ". Install with: install.packages(c(",
      paste(sprintf("\"%s\"", missing_packages), collapse = ", "),
      "))"
    )
  )
}

library(data.table)
library(dplyr)
library(DT)
library(ggplot2)
library(knitr)
library(skimr)
library(scales)
library(stringr)
library(tidyr)

options(
  scipen = 999,
  datatable.print.nrows = 50,
  DT.options = list(pageLength = 25, scrollX = TRUE)
)

`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0 || all(is.na(x))) y else x
}

normalize_names <- function(dt) {
  data.table::setnames(dt, names(dt), tolower(names(dt)))
  dt
}

clean_character <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

clean_boolish <- function(x) {
  x <- clean_character(x)
  upper <- toupper(x)
  dplyr::case_when(
    upper %in% c("TRUE", "T", "YES", "Y", "1") ~ "TRUE",
    upper %in% c("FALSE", "F", "NO", "N", "0") ~ "FALSE",
    is.na(upper) ~ NA_character_,
    TRUE ~ x
  )
}

coerce_numeric <- function(dt, cols) {
  for (col in intersect(cols, names(dt))) {
    dt[, (col) := suppressWarnings(as.numeric(get(col)))]
  }
  invisible(dt)
}

current_rmd_dir <- function() {
  input <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
  if (is.null(input) || !nzchar(input)) {
    return(getwd())
  }
  dirname(normalizePath(input, winslash = "/", mustWork = FALSE))
}

resolve_path <- function(path, must_work = TRUE) {
  candidates <- c(
    path,
    file.path(current_rmd_dir(), path),
    file.path(current_rmd_dir(), "..", path)
  )
  existing <- candidates[file.exists(candidates)]
  if (length(existing) > 0) {
    return(normalizePath(existing[1], winslash = "/", mustWork = TRUE))
  }
  if (isTRUE(must_work)) {
    stop("Path does not exist: ", path)
  }
  normalizePath(candidates[1], winslash = "/", mustWork = FALSE)
}

make_output_dir <- function(base_dir) {
  out_dir <- file.path(resolve_path(base_dir, must_work = FALSE), format(Sys.time(), "%Y%m%d_%H%M%S"))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  out_dir
}

write_output <- function(x, name) {
  data.table::fwrite(x, file.path(output_dir, paste0(name, ".csv")), na = "")
  invisible(x)
}

display_table <- function(x, caption = NULL, page_length = 25) {
  if (nrow(x) > params$max_interactive_rows) {
    x <- head(x, params$max_interactive_rows)
    caption <- paste0(
      caption %||% "Table",
      " (showing first ",
      scales::comma(params$max_interactive_rows),
      " rows; full table written to CSV)"
    )
  }
  DT::datatable(
    as.data.frame(x),
    caption = caption,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = page_length, scrollX = TRUE)
  )
}

historical_input_dir <- resolve_path(params$input_dir, must_work = TRUE)
holdout_dir <- file.path(historical_input_dir, "Printing Schedule (On Sale July 2024)")
output_dir <- make_output_dir(params$output_subdir)
set.seed(42)
```

# Scope

This notebook intentionally reads only the historical training files directly under `r historical_input_dir`.
It does not read `r holdout_dir`, which is the July 2024 holdout folder.

The analysis grain is one historical `STORE_ID x PRODUCT_ID` fact row after joining product, store, and optionally demographic attributes. Tables are grouped by `TYPE`, with the practical focus on `Weeklies` and `SIP`.

```{r paths}
paths <- list(
  fact = file.path(historical_input_dir, "FACT_TABLE.csv"),
  product = file.path(historical_input_dir, "DIM_PRODUCT.csv"),
  store = file.path(historical_input_dir, "DIM_STORE.csv"),
  demographics = file.path(historical_input_dir, "DEMOGRAPHICS.csv")
)

path_audit <- data.table(
  source = names(paths),
  path = unlist(paths, use.names = FALSE),
  exists = file.exists(unlist(paths, use.names = FALSE))
)
path_audit[, july_holdout_path := grepl("Printing Schedule \\(On Sale July 2024\\)", path)]
write_output(path_audit, "path_audit")
display_table(path_audit, "Historical input paths used by this EDA")
```

# Load And Join Data

```{r load-data}
fact <- fread(paths$fact, showProgress = TRUE) |> normalize_names()
product <- fread(paths$product, showProgress = TRUE) |> normalize_names()
store <- fread(paths$store, showProgress = TRUE) |> normalize_names()

demographics <- NULL
if (isTRUE(params$include_demographics) && file.exists(paths$demographics)) {
  demographics <- fread(paths$demographics, showProgress = TRUE) |> normalize_names()
}

coerce_numeric(fact, c("store_id", "product_id", "soldqty", "drawqty"))
coerce_numeric(product, c("product_id", "price"))
coerce_numeric(store, c("store_id", "facings", "pockets"))

product[, `:=`(
  barcode = clean_character(barcode),
  title = clean_character(title),
  type = clean_character(type),
  segment = clean_character(segment),
  subsegment = clean_character(subsegment),
  frequency = clean_character(frequency),
  onsaledate = as.IDate(onsaledate),
  offsaledate = as.IDate(offsaledate)
)]

store[, `:=`(
  postal_code = clean_character(postal_code),
  store_chain = clean_character(store_chain),
  region = clean_character(region),
  classoftrade = clean_character(classoftrade),
  merchandised = clean_boolish(merchandised)
)]

if (!is.null(demographics)) {
  demographics[, postal_code := clean_character(postal_code)]
  demographic_numeric_cols <- setdiff(names(demographics), "postal_code")
  coerce_numeric(demographics, demographic_numeric_cols)

  demographics <- demographics[
    ,
    lapply(.SD, function(x) {
      if (is.numeric(x)) {
        if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)
      } else {
        x[which.max(!is.na(x))]
      }
    }),
    by = postal_code
  ]
}

setkey(product, product_id)
setkey(store, store_id)
setkey(fact, product_id)
analysis <- product[fact, nomatch = 0]
setkey(analysis, store_id)
analysis <- store[analysis, nomatch = 0]

if (!is.null(demographics)) {
  setkey(demographics, postal_code)
  setkey(analysis, postal_code)
  analysis <- demographics[analysis]
}

analysis[, `:=`(
  type = fifelse(is.na(type), "UNKNOWN", type),
  issue_length_days = as.numeric(offsaledate - onsaledate),
  onsale_month = as.integer(format(onsaledate, "%m")),
  onsale_week = as.integer(strftime(onsaledate, format = "%V")),
  sellthrough = fifelse(drawqty > 0, soldqty / drawqty, NA_real_),
  oversupply_units = pmax(drawqty - pmax(soldqty, 0), 0),
  stockout_proxy_flag = drawqty > 0 & soldqty == drawqty,
  negative_soldqty_flag = soldqty < 0,
  zero_soldqty_flag = soldqty == 0
)]

target_types <- c("Weeklies", "SIP")
analysis <- analysis[type %in% target_types]
analysis[, type := factor(type, levels = target_types)]
```

```{r data-audit}
data_audit <- rbindlist(list(
  data.table(table = "FACT_TABLE", rows = nrow(fact), columns = ncol(fact)),
  data.table(table = "DIM_PRODUCT", rows = nrow(product), columns = ncol(product)),
  data.table(table = "DIM_STORE", rows = nrow(store), columns = ncol(store)),
  data.table(table = "DEMOGRAPHICS", rows = ifelse(is.null(demographics), 0L, nrow(demographics)), columns = ifelse(is.null(demographics), 0L, ncol(demographics))),
  data.table(table = "ANALYSIS_PANEL_WEEKLIES_SIP", rows = nrow(analysis), columns = ncol(analysis))
))

type_overview <- analysis[
  ,
  .(
    fact_rows = .N,
    stores = uniqueN(store_id),
    products = uniqueN(product_id),
    titles = uniqueN(title, na.rm = TRUE),
    segments = uniqueN(segment, na.rm = TRUE),
    subsegments = uniqueN(subsegment, na.rm = TRUE),
    first_onsale = min(onsaledate, na.rm = TRUE),
    last_onsale = max(onsaledate, na.rm = TRUE),
    soldqty_sum = sum(soldqty, na.rm = TRUE),
    drawqty_sum = sum(drawqty, na.rm = TRUE),
    aggregate_sellthrough = sum(soldqty, na.rm = TRUE) / sum(drawqty, na.rm = TRUE),
    row_avg_sellthrough = mean(sellthrough, na.rm = TRUE),
    zero_sales_rate = mean(zero_soldqty_flag, na.rm = TRUE),
    negative_sales_rate = mean(negative_soldqty_flag, na.rm = TRUE),
    stockout_proxy_rate = mean(stockout_proxy_flag, na.rm = TRUE)
  ),
  by = type
]

write_output(data_audit, "data_audit")
write_output(type_overview, "type_overview")

display_table(data_audit, "Loaded data audit")
display_table(type_overview, "Weeklies vs SIP row-grain overview")
```

# Feature Lists

```{r feature-lists}
base_categorical_features <- c(
  "title",
  "barcode",
  "segment",
  "subsegment",
  "frequency",
  "store_chain",
  "region",
  "classoftrade",
  "merchandised",
  "postal_code"
)

engineered_categorical_features <- c(
  "onsale_month",
  "onsale_week"
)

base_numeric_features <- c(
  "soldqty",
  "drawqty",
  "sellthrough",
  "oversupply_units",
  "price",
  "issue_length_days",
  "facings",
  "pockets"
)

demographic_numeric_features <- if (!is.null(demographics)) {
  setdiff(names(demographics), "postal_code")
} else {
  character(0)
}

categorical_features <- intersect(
  c(base_categorical_features, engineered_categorical_features),
  names(analysis)
)

numeric_features <- intersect(
  c(base_numeric_features, demographic_numeric_features),
  names(analysis)
)

feature_inventory <- data.table(
  feature = c(categorical_features, numeric_features),
  feature_role = c(rep("categorical", length(categorical_features)), rep("numeric", length(numeric_features)))
)

write_output(feature_inventory, "feature_inventory")
display_table(feature_inventory, "Features profiled in this notebook")
```

# Categorical Values

The table in this section is the full inventory of categorical values by product type and feature. It is also written to `categorical_value_inventory.csv` in the output directory.

```{r categorical-inventory}
categorical_summary_one <- function(dt, feature) {
  value <- as.character(dt[[feature]])
  value[is.na(value) | trimws(value) == ""] <- "(missing)"
  tmp <- data.table(
    type = dt$type,
    feature = feature,
    value = value,
    product_id = dt$product_id,
    store_id = dt$store_id,
    soldqty = dt$soldqty,
    drawqty = dt$drawqty,
    sellthrough = dt$sellthrough,
    zero_soldqty_flag = dt$zero_soldqty_flag,
    stockout_proxy_flag = dt$stockout_proxy_flag
  )

  tmp[
    ,
    .(
      row_count = .N,
      row_share_within_type_feature = .N / .N,
      product_count = uniqueN(product_id),
      store_count = uniqueN(store_id),
      soldqty_sum = sum(soldqty, na.rm = TRUE),
      drawqty_sum = sum(drawqty, na.rm = TRUE),
      aggregate_sellthrough = sum(soldqty, na.rm = TRUE) / sum(drawqty, na.rm = TRUE),
      row_avg_sellthrough = mean(sellthrough, na.rm = TRUE),
      zero_sales_rate = mean(zero_soldqty_flag, na.rm = TRUE),
      stockout_proxy_rate = mean(stockout_proxy_flag, na.rm = TRUE)
    ),
    by = .(type, feature, value)
  ][
    ,
    row_share_within_type_feature := row_count / sum(row_count),
    by = .(type, feature)
  ][order(type, feature, -row_count, value)]
}

categorical_value_inventory <- rbindlist(
  lapply(categorical_features, function(feature) categorical_summary_one(analysis, feature)),
  use.names = TRUE,
  fill = TRUE
)

categorical_cardinality <- categorical_value_inventory[
  ,
  .(
    distinct_values = .N,
    missing_values_present = any(value == "(missing)"),
    largest_value_row_share = max(row_share_within_type_feature, na.rm = TRUE),
    top_value = value[which.max(row_count)]
  ),
  by = .(type, feature)
][order(type, -distinct_values)]

write_output(categorical_cardinality, "categorical_cardinality")
write_output(categorical_value_inventory, "categorical_value_inventory")

display_table(categorical_cardinality, "Categorical feature cardinality by Weeklies and SIP")
display_table(categorical_value_inventory, "Full categorical value inventory by Weeklies and SIP")
```

```{r categorical-preview}
categorical_preview <- categorical_value_inventory[
  ,
  head(.SD, params$top_n_preview),
  by = .(type, feature)
]

display_table(categorical_preview, paste0("Top ", params$top_n_preview, " categorical values per feature and type"))
```

```{r categorical-cardinality-plot}
ggplot(categorical_cardinality, aes(x = reorder(feature, distinct_values), y = distinct_values, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Categorical Cardinality By Product Type",
    x = NULL,
    y = "Distinct values"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r top-categorical-values-plot}
plot_cats <- categorical_value_inventory[
  feature %in% c("segment", "subsegment", "frequency", "store_chain", "region", "classoftrade", "merchandised"),
  head(.SD, params$top_n_plot),
  by = .(type, feature)
]

ggplot(plot_cats, aes(x = reorder(value, row_count), y = row_count, fill = type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_grid(type ~ feature, scales = "free_y", space = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Most Common Categorical Values",
    x = NULL,
    y = "Historical row count"
  ) +
  theme_minimal(base_size = 10)
```

# Numerical Feature Ranges

The main numeric table gives range and distribution statistics separately for Weeklies and SIP. It is also written to `numeric_feature_ranges.csv`.

```{r numeric-ranges}
numeric_summary_one <- function(dt, feature) {
  x <- suppressWarnings(as.numeric(dt[[feature]]))
  data.table(type = dt$type, value = x)[
    ,
    {
      clean <- value[!is.na(value)]
      qs <- if (length(clean) > 0) {
        as.numeric(stats::quantile(clean, probs = c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99), na.rm = TRUE, names = FALSE, type = 7))
      } else {
        rep(NA_real_, 7)
      }
      .(
        feature = feature,
        row_count = .N,
        non_missing_count = sum(!is.na(value)),
        missing_count = sum(is.na(value)),
        missing_rate = mean(is.na(value)),
        distinct_values = uniqueN(value, na.rm = TRUE),
        min = ifelse(length(clean) > 0, min(clean), NA_real_),
        p01 = qs[1],
        p05 = qs[2],
        q25 = qs[3],
        median = qs[4],
        mean = ifelse(length(clean) > 0, mean(clean), NA_real_),
        q75 = qs[5],
        p95 = qs[6],
        p99 = qs[7],
        max = ifelse(length(clean) > 0, max(clean), NA_real_),
        sd = ifelse(length(clean) > 1, stats::sd(clean), NA_real_),
        negative_count = sum(value < 0, na.rm = TRUE),
        zero_count = sum(value == 0, na.rm = TRUE)
      )
    },
    by = type
  ][, .(
    type,
    feature,
    row_count,
    non_missing_count,
    missing_count,
    missing_rate,
    distinct_values,
    min,
    p01,
    p05,
    q25,
    median,
    mean,
    q75,
    p95,
    p99,
    max,
    sd,
    negative_count,
    zero_count
  )]
}

numeric_feature_ranges <- rbindlist(
  lapply(numeric_features, function(feature) numeric_summary_one(analysis, feature)),
  use.names = TRUE,
  fill = TRUE
)[order(type, feature)]

write_output(numeric_feature_ranges, "numeric_feature_ranges")
display_table(numeric_feature_ranges, "Numeric feature ranges by Weeklies and SIP")
```

```{r numeric-ranges-plot}
range_plot <- numeric_feature_ranges[
  non_missing_count > 0 & is.finite(p01) & is.finite(p99) & p01 != p99
]

ggplot(range_plot, aes(x = p01, xend = p99, y = reorder(feature, p99), color = type)) +
  geom_segment(position = position_dodge(width = 0.7), linewidth = 1.2) +
  geom_point(aes(x = median), position = position_dodge(width = 0.7), size = 2) +
  facet_wrap(~ type, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Numeric Feature Robust Ranges",
    subtitle = "Segment spans p01 to p99; point marks median",
    x = "Value",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r numeric-distribution-sample}
sample_n <- min(params$row_sample_n, nrow(analysis))
analysis_sample <- analysis[sample(.N, sample_n)]

numeric_plot_features <- intersect(
  c("soldqty", "drawqty", "sellthrough", "oversupply_units", "price", "issue_length_days", "facings", "pockets", "population"),
  numeric_features
)

numeric_sample_long <- melt(
  analysis_sample,
  id.vars = "type",
  measure.vars = numeric_plot_features,
  variable.name = "feature",
  value.name = "value"
)
numeric_sample_long <- numeric_sample_long[!is.na(value) & is.finite(value)]

ggplot(numeric_sample_long, aes(x = type, y = value, fill = type)) +
  geom_boxplot(outlier.alpha = 0.08) +
  facet_wrap(~ feature, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Sampled Numeric Distributions By Product Type",
    subtitle = paste("Sampled rows:", comma(sample_n)),
    x = NULL,
    y = "Value"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

# Missingness

```{r missingness}
missingness_by_type <- rbindlist(lapply(c(categorical_features, numeric_features), function(feature) {
  x <- analysis[[feature]]
  data.table(type = analysis$type, feature = feature, feature_role = ifelse(feature %in% categorical_features, "categorical", "numeric"), missing = is.na(x) | (is.character(x) & trimws(x) == ""))[
    ,
    .(
      row_count = .N,
      missing_count = sum(missing),
      missing_rate = mean(missing)
    ),
    by = .(type, feature, feature_role)
  ]
}))

write_output(missingness_by_type, "missingness_by_type")
display_table(missingness_by_type[order(type, -missing_rate, feature)], "Missingness by feature and product type")
```

## Skimr By Product Type

This section adds standard `skimr` summaries for each product type separately so the raw missing counts are visible in one familiar table per group.

```{r skimr-by-type}
skim_columns <- setdiff(names(analysis), "type")

skim_by_type <- lapply(target_types, function(current_type) {
  skim_tbl <- analysis[type == current_type, ..skim_columns] |>
    as.data.frame() |>
    skimr::skim_without_charts()

  skim_tbl <- as.data.table(skim_tbl)
  skim_tbl[, type := current_type]
  data.table::setcolorder(skim_tbl, c("type", setdiff(names(skim_tbl), "type")))
  skim_tbl
})

skim_weeklies <- skim_by_type[[which(target_types == "Weeklies")]]
skim_sip <- skim_by_type[[which(target_types == "SIP")]]
skim_combined <- rbindlist(skim_by_type, use.names = TRUE, fill = TRUE)

write_output(skim_weeklies, "skimr_weeklies")
write_output(skim_sip, "skimr_sip")
write_output(skim_combined, "skimr_by_type")

display_table(skim_weeklies, "skimr summary for Weeklies")
display_table(skim_sip, "skimr summary for SIP")
```

```{r missingness-plot}
ggplot(missingness_by_type, aes(x = reorder(feature, missing_rate), y = missing_rate, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ feature_role, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Feature Missingness By Product Type",
    x = NULL,
    y = "Missing row share"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

# Product And Store Coverage

```{r product-store-coverage}
product_coverage <- analysis[
  ,
  .(
    store_rows = .N,
    stores = uniqueN(store_id),
    soldqty_sum = sum(soldqty, na.rm = TRUE),
    drawqty_sum = sum(drawqty, na.rm = TRUE),
    aggregate_sellthrough = sum(soldqty, na.rm = TRUE) / sum(drawqty, na.rm = TRUE),
    title = first(title),
    barcode = first(barcode),
    segment = first(segment),
    subsegment = first(subsegment),
    frequency = first(frequency),
    onsaledate = first(onsaledate),
    offsaledate = first(offsaledate),
    price = first(price)
  ),
  by = .(type, product_id)
][order(type, -store_rows)]

store_coverage <- analysis[
  ,
  .(
    fact_rows = .N,
    products = uniqueN(product_id),
    titles = uniqueN(title, na.rm = TRUE),
    soldqty_sum = sum(soldqty, na.rm = TRUE),
    drawqty_sum = sum(drawqty, na.rm = TRUE),
    aggregate_sellthrough = sum(soldqty, na.rm = TRUE) / sum(drawqty, na.rm = TRUE),
    store_chain = first(store_chain),
    region = first(region),
    classoftrade = first(classoftrade),
    merchandised = first(merchandised),
    facings = first(facings),
    pockets = first(pockets),
    postal_code = first(postal_code)
  ),
  by = .(type, store_id)
][order(type, -fact_rows)]

write_output(product_coverage, "product_coverage")
write_output(store_coverage, "store_coverage")

display_table(product_coverage, "Product-level coverage and outcomes")
display_table(store_coverage, "Store-level coverage and outcomes")
```

# Outputs

```{r output-manifest}
output_manifest <- data.table(
  output_dir = output_dir,
  file = basename(list.files(output_dir, full.names = FALSE)),
  path = list.files(output_dir, full.names = TRUE)
)

display_table(output_manifest, "CSV outputs written by this notebook")
cat("Output directory:", output_dir, "\n")
```
